Appendix C: Latent Class Growth Analysis

### 2023-01-24 appendix_b.Rmd
revert MGV previous commit

---
title             : "Appendix B: Exploratory LPA for Ocean Microplastics"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
bibliography      : "r-references.bib"
floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
editor_options: 
  markdown: 
    wrap: sentence
---

@@

This is an example of exploratory latent class analysis (LCA) with continuous indicators, otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using `tidySEM`.
The present example uses data collected by Alkema as part of a study on ocean microplastics.
The purpose of this study was to provide a more nuanced model for the distribution of different sizes of ocean microplastics than the commonly used normal distribution.
To this end, a mixture of normals was used.
Since there is no theoretical reason to expect a certain number of classes, this is an exploratory LCA.
To view its documentation, run the command `?tidySEM::alkema_microplastics` in the R console.
The original analyses are available at <https://github.com/cjvanlissa/lise_microplastics>; in this vignette, we take a different approach to the analysis to showcase other possibilities.

@@

```{r, echo = TRUE, eval = FALSE}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
```

```{r, echo = FALSE, eval = run_everything}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
p <- ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
ggsave("appendices/plot_gmm_desc.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```

```{r figdesc, echo = FALSE, eval = pkgdown}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
knitr::include_graphics("appendices/plot_gmm_desc.pdf")
```

@@

```{r, echo = TRUE, eval = FALSE}
df_plot$Value <- log(df_plot$Value)
ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
```

```{r, echo = FALSE, eval = run_everything}
df_plot$Value <- log(df_plot$Value)
p <- ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
ggsave("appendices/plot_gmm_desc_log.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```

```{r figdesclog, echo = FALSE, eval = pkgdown}
knitr::include_graphics("appendices/plot_gmm_desc_log.pdf")
```

@@

```{r, eval = FALSE, echo = TRUE}
df <- reshape(df_plot, direction = "wide", v.names = "Value")[, -1]
names(df) <- gsub("Value.", "", names(df), fixed = TRUE)
ggplot(df, aes(x = length, y = width)) +
  geom_point(alpha = .1) +
  theme_bw()
```

```{r, eval = run_everything, echo = FALSE}
df <- reshape(df_plot, direction = "wide", v.names = "Value")[, -1]
names(df) <- gsub("Value.", "", names(df), fixed = TRUE)
p <- ggplot(df, aes(x = length, y = width)) +
  geom_point(alpha = .1) +
  theme_bw()
ggsave("appendices/plot_gmm_scatter.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```

```{r figscatter, eval = pkgdown, echo = FALSE}
knitr::include_graphics("appendices/plot_gmm_scatter.pdf")
```

<!-- As evident from the scatterplot, there is a very strong linear tend in the data. -->

<!-- This indicates that fragments tend to be coextensive (when length goes up, width goes up). -->

<!-- We can analyze these data in their original dimensions (length and width). -->

<!-- Alternatively, we can use PCA to rotate the data such that the first dimension can be interpreted as "size", -->

<!-- and the second dimension can be interpreted as "shape": -->

@@

```{r fitlca, eval = FALSE, echo = TRUE}
set.seed(123)
res <- mx_profiles(data = df,
                   classes = 1:4,
                   variances = c("equal", "varying"),
                   covariances = c("zero", "equal",
                                   "varying"),
                   expand_grid = TRUE)
saveRDS(res, "res_gmm.RData")
```

```{r eval = run_everything, echo = FALSE}
set.seed(123)
res <- mx_profiles(data = df,
                   classes = 1:4,
                   variances = c("equal", "varying"),
                   covariances = c("zero", "equal",
                                   "varying"),
                   expand_grid = TRUE)
saveRDS(res, "appendices/res_gmm.RData")
fit <- table_fit(res)
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
saveRDS(fit, "appendices/gmm_tabfit.RData")
#"Warning: In model 'mix4' Optimizer returned a non-zero status code 6. The model does not satisfy the first-order optimality conditions to the required accuracy, and no improved point for the merit function could be found during the final linesearch (Mx status RED)"
```

```{r eval = TRUE, echo = FALSE}
fit <- readRDS("appendices/gmm_tabfit.RData")
```

@@

<!-- This suggests a potential problem with with model convergence. -->

<!-- To try and get this model to converge, it is possible to use the function `mxTryHard()`, which permutes the starting values and tries to find a better solution. -->

<!-- To do this, run `res[[24]] <- mxTryHard(res[[24]])`. -->

<!-- However, at present, we will skip this step, -->

<!-- because there are also other indications that this 4-class model is overfitted (see below). -->

@@

```{r echo = TRUE, eval = FALSE}
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
fit[ , c("Name", "LL", "Parameters", "par_ratio",
         "BIC", "Entropy",
         "prob_min", "prob_max", 
         "n_min", "n_max",
         "lmr_p")]
```

```{r tabfit, echo = FALSE, eval = TRUE}
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
tbl <- fit[ , c("Name", "LL", "Parameters", "par_ratio",
         "BIC", "Entropy",
         "prob_min",
         "n_min",
         "lmr_p")]
names(tbl) <- c("Name", "LL", "p", "par_ratio",
         "BIC", "Ent.",
         "p_min",
         "n_min",
         "lmr_p")
papaja::apa_table(tbl, caption = "Model fit table. Some columns removed to fit page.")
```

@@

As mentioned before, we drop models with \< 10% of cases in the smallest class:

@@

```{r echo = TRUE, eval = FALSE}
plot(fit) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r echo = FALSE, eval = run_everything}
p <- plot(fit) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("appendices/gmm_plotfit.pdf", p, device = "pdf", width = 100, height = 100, units = "mm", scale = 1)
```

```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
knitr::include_graphics("appendices/gmm_plotfit.pdf")
```

@@

Out of the 16 models that remain after removing those with \< 10% of cases in the smallest class, one model stands out: The 2-class model with free (co)variances.
<!-- The function `ic_weights(fit)` allows us to compute IC weights for all models in the set; --> <!-- it prefers the most complex model out of the 16 models with a posterior model probability of nearly 100%. --> We thus select this as our final model.

@@

```{r echo = TRUE, eval = FALSE}
res_bic <- res[["free var, free cov 2"]]
cp <- class_prob(res_bic)
results <- table_results(res_bic, columns = c("label", "est", "std_est"))
results
```

```{r echo = FALSE, eval = run_everything}
res_bic <- res[["free var, free cov 2"]]
cp <- class_prob(res_bic)
results <- table_results(res_bic, columns = c("label", "est", "est_std"))
saveRDS(results, "appendices/gmm_results.RData")
```

```{r eval = TRUE, echo = FALSE}
results <- readRDS("appendices/gmm_results.RData")
papaja::apa_table(results, digits = 2, caption = "Results of a 2-class model with free (co)variances")
```

@@

```{r echo = TRUE, eval = FALSE}
plot_bivariate(res_bic)
```

```{r echo = FALSE, eval = run_everything}
p <- plot_bivariate(res_bic)
ggsave("appendices/gmm_bivariate_bic.pdf", p, device = "pdf", width = 210, height = 100, units = "mm", scale = 1.5)
```

```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
knitr::include_graphics("appendices/gmm_bivariate_bic.pdf")
```

@@

```{r echo = TRUE, eval=FALSE}
df_pt <- mx_dummies(df_analyze$poly_type)
aux_pt <- BCH(res_bic, model = "poly_typeOther | t1
                                poly_typePE | t1
                                poly_typePP | t1", data = df_pt)
aux_pt <- mxTryHardOrdinal(aux_pt)
```

```{r echo = FALSE, eval=run_everything}
df_pt <- mx_dummies(df_analyze$poly_type)
aux_pt <- BCH(res_final, model = "poly_typeOther | t1
                                  poly_typePE | t1
                                  poly_typePP | t1", data = df_pt)
aux_pt <- mxTryHardOrdinal(aux_pt)
saveRDS(aux_pt, "appendices/gmm_aux_pt.RData")
```

```{r echo = FALSE, eval=FALSE}
aux_pt <- readRDS("appendices/gmm_aux_pt.RData")
```

>>>>

---
title             : "Appendix B: Exploratory LPA for Ocean Microplastics"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"

bibliography      : "r-references.bib"

floatsintext      : yes
linenumbers       : no
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "man"
output            : papaja::apa6_pdf
---

@@

This is an example of exploratory latent class analysis (LCA) with continuous indicators,
otherwise known as latent profile analysis (LPA) or finite Gaussian mixture modeling, using `tidySEM`.
The present example uses data collected by Alkema as part of a study on ocean microplastics.
The purpose of this study was to provide a more nuanced model for the distribution of different sizes of ocean microplastics than the commonly used normal distribution.
To this end, a mixture of normals was used.
Since there is no theoretical reason to expect a certain number of classes, this is an exploratory LCA.
To view its documentation,
run the command `?tidySEM::alkema_microplastics` in the R console.
The original analyses are available at https://github.com/cjvanlissa/lise_microplastics;
in this vignette, we take a different approach to the analysis to showcase other possibilities.

@@

```{r, echo = TRUE, eval = FALSE}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
```
```{r, echo = FALSE, eval = run_everything}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
p <- ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
ggsave("appendices/plot_gmm_desc.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```
```{r figdesc, echo = FALSE, eval = pkgdown}
df_plot <- df
names(df_plot) <- paste0("Value.", names(df_plot))
df_plot <- reshape(df_plot, varying = names(df_plot), direction = "long",
                   timevar = "Variable")
knitr::include_graphics("appendices/plot_gmm_desc.pdf")
```

@@

```{r, echo = TRUE, eval = FALSE}
df_plot$Value <- log(df_plot$Value)
ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
```
```{r, echo = FALSE, eval = run_everything}
df_plot$Value <- log(df_plot$Value)
p <- ggplot(df_plot, aes(x = Value)) +
  geom_density() +
  facet_wrap(~Variable)+
  theme_bw()
ggsave("appendices/plot_gmm_desc_log.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```
```{r figdesclog, echo = FALSE, eval = pkgdown}
knitr::include_graphics("appendices/plot_gmm_desc_log.pdf")
```

@@

```{r, eval = FALSE, echo = TRUE}
df <- reshape(df_plot, direction = "wide", v.names = "Value")[, -1]
names(df) <- gsub("Value.", "", names(df), fixed = TRUE)
ggplot(df, aes(x = length, y = width)) +
  geom_point(alpha = .1) +
  theme_bw()
```
```{r, eval = run_everything, echo = FALSE}
df <- reshape(df_plot, direction = "wide", v.names = "Value")[, -1]
names(df) <- gsub("Value.", "", names(df), fixed = TRUE)
p <- ggplot(df, aes(x = length, y = width)) +
  geom_point(alpha = .1) +
  theme_bw()
ggsave("appendices/plot_gmm_scatter.pdf", p, device = "pdf", width = 100, height = 100, units = "mm")
```
```{r figscatter, eval = pkgdown, echo = FALSE}
knitr::include_graphics("appendices/plot_gmm_scatter.pdf")
```

<!-- As evident from the scatterplot, there is a very strong linear tend in the data. -->
<!-- This indicates that fragments tend to be coextensive (when length goes up, width goes up). -->
<!-- We can analyze these data in their original dimensions (length and width). -->
<!-- Alternatively, we can use PCA to rotate the data such that the first dimension can be interpreted as "size", -->
<!-- and the second dimension can be interpreted as "shape": -->

@@

```{r fitlca, eval = FALSE, echo = TRUE}
set.seed(123)
res <- mx_profiles(data = df,
                   classes = 1:4,
                   variances = c("equal", "varying"),
                   covariances = c("zero", "equal",
                                   "varying"),
                   expand_grid = TRUE)
saveRDS(res, "res_gmm.RData")
```
```{r eval = run_everything, echo = FALSE}
set.seed(123)
res <- mx_profiles(data = df,
                   classes = 1:4,
                   variances = c("equal", "varying"),
                   covariances = c("zero", "equal",
                                   "varying"),
                   expand_grid = TRUE)
saveRDS(res, "appendices/res_gmm.RData")
fit <- table_fit(res)
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
saveRDS(fit, "appendices/gmm_tabfit.RData")
#"Warning: In model 'mix4' Optimizer returned a non-zero status code 6. The model does not satisfy the first-order optimality conditions to the required accuracy, and no improved point for the merit function could be found during the final linesearch (Mx status RED)"
```
```{r eval = TRUE, echo = FALSE}
fit <- readRDS("appendices/gmm_tabfit.RData")
```

@@

<!-- This suggests a potential problem with with model convergence. -->
<!-- To try and get this model to converge, it is possible to use the function `mxTryHard()`, which permutes the starting values and tries to find a better solution. -->
<!-- To do this, run `res[[24]] <- mxTryHard(res[[24]])`. -->
<!-- However, at present, we will skip this step, -->
<!-- because there are also other indications that this 4-class model is overfitted (see below). -->

@@

```{r echo = TRUE, eval = FALSE}
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
fit[ , c("Name", "LL", "Parameters", "par_ratio",
         "BIC", "Entropy",
         "prob_min", "prob_max", 
         "n_min", "n_max",
         "lmr_p")]
```
```{r tabfit, echo = FALSE, eval = TRUE}
fit$par_ratio <- (5605*fit$n_min) / (fit$Parameters/fit$Classes)
tbl <- fit[ , c("Name", "LL", "Parameters", "par_ratio",
         "BIC", "Entropy",
         "prob_min",
         "n_min",
         "lmr_p")]
names(tbl) <- c("Name", "LL", "p", "par_ratio",
         "BIC", "Ent.",
         "p_min",
         "n_min",
         "lmr_p")
papaja::apa_table(tbl, caption = "Model fit table. Some columns removed to fit page.")
```

@@

As mentioned before, we drop models with < 10% of cases in the smallest class:

@@

```{r echo = TRUE, eval = FALSE}
plot(fit) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
```{r echo = FALSE, eval = run_everything}
p <- plot(fit) + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
ggsave("appendices/gmm_plotfit.pdf", p, device = "pdf", width = 100, height = 100, units = "mm", scale = 1)
```
```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
knitr::include_graphics("appendices/gmm_plotfit.pdf")
```

@@

Out of the 16 models that remain after removing those with < 10% of cases in the smallest class,
one model stands out:
The 2-class model with free (co)variances.
<!-- The function `ic_weights(fit)` allows us to compute IC weights for all models in the set; -->
<!-- it prefers the most complex model out of the 16 models with a posterior model probability of nearly 100%. -->
We thus select this as our final model.

@@

```{r echo = TRUE, eval = FALSE}
res_bic <- res[["free var, free cov 2"]]
cp <- class_prob(res_bic)
results <- table_results(res_bic, columns = c("label", "est", "std_est"))
results
```
```{r echo = FALSE, eval = run_everything}
res_bic <- res[["free var, free cov 2"]]
cp <- class_prob(res_bic)
results <- table_results(res_bic, columns = c("label", "est", "est_std"))
saveRDS(results, "appendices/gmm_results.RData")
```
```{r eval = TRUE, echo = FALSE}
results <- readRDS("appendices/gmm_results.RData")
papaja::apa_table(results, digits = 2, caption = "Results of a 2-class model with free (co)variances")
```

@@

```{r echo = TRUE, eval = FALSE}
plot_bivariate(res_bic)
```
```{r echo = FALSE, eval = run_everything}
p <- plot_bivariate(res_bic)
ggsave("appendices/gmm_bivariate_bic.pdf", p, device = "pdf", width = 210, height = 100, units = "mm", scale = 1.5)
```
```{r echo = FALSE, eval = pkgdown, fig.cap="Bivariate profile plot", out.width="100%"}
knitr::include_graphics("appendices/gmm_bivariate_bic.pdf")
```

@@

```{r echo = TRUE, eval=FALSE}
df_pt <- mx_dummies(df_analyze$poly_type)
aux_pt <- BCH(res_bic, model = "poly_typeOther | t1
                                poly_typePE | t1
                                poly_typePP | t1", data = df_pt)
aux_pt <- mxTryHardOrdinal(aux_pt)
```
```{r echo = FALSE, eval=run_everything}
df_pt <- mx_dummies(df_analyze$poly_type)
aux_pt <- BCH(res_final, model = "poly_typeOther | t1
                                  poly_typePE | t1
                                  poly_typePP | t1", data = df_pt)
aux_pt <- mxTryHardOrdinal(aux_pt)
saveRDS(aux_pt, "appendices/gmm_aux_pt.RData")
```
```{r echo = FALSE, eval=FALSE}
aux_pt <- readRDS("appendices/gmm_aux_pt.RData")
```
